---
title: "Run eQTM analysis for the NCI-Maryland cohort on expression and methylation residuals"
output: html_document
date: "2025-01-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(stringr)
```

# Load data
Probe-gene pairs, along with genomic coordinates
```{r}
probe_gene_pairs <- read_csv("AFR-probe-gene-pairs.csv")
View(probe_gene_pairs)
```
Gene expression residuals
```{r}
expression <- read_csv("residuals_DESeq2expression_tumors_GeneSymbol.csv")
head(expression)
```
Beta value residuals
```{r}
methylation <- read_csv("residuals_methylation_tumors.csv")
head(methylation)
```
Keep same patients between methylation and expression matrix
```{r}
colnames(methylation)[1] <- "probe"
methylation <- methylation[,c("probe",intersect(colnames(methylation), colnames(expression)))]
```

Create appropriate format for beta matrix
```{r}
methylation <- as.data.frame(methylation)
row.names(methylation) <- methylation[,1]
methylation <- methylation[,-1]
```
Preprocess and format expression data
```{r}
expression <- as.data.frame(expression)

row.names(expression) <- expression[,1]
expression <- expression[,-1]
```

From file with candidate eQTMs, keep only probe-gene pairs
```{r}
pairs <- probe_gene_pairs[,c(6,12)]
colnames(pairs) <- c("probe", "gene_symbol")
head(pairs)
```

Make sure all samples are Female
```{r}
meth_metadata <- read_csv("meth_metadata.csv")

sel <- meth_metadata[,c("title","Sex:ch1")]

sel$accession <- str_sub(sel$title,32,36)
sel$tissue_type <- word(sel$title, -1) 

sel2 <- sel[which(sel$accession %in% colnames(expression) & sel$tissue_type == "tumor"),]
table(sel2$`Sex:ch1`)
```

Remove male samples from expression and methylation df
```{r}
expression2 <- expression[,-(which(colnames(expression) %in% sel2$accession[which(sel2$`Sex:ch1` == "M")]))]
```
Remove male samples from methylation df as well
```{r}
methylation2 <- methylation[,colnames(expression2)]
all.equal(colnames(expression2), colnames(methylation2)) # sanity check for sample order
```

## Run eQTM analysis
Create dataframe to store the results
```{r}
overall_results <- data.frame(probe = pairs$probe, gene=pairs$gene_symbol)
```

Write loop
```{r}
for (i in 1:dim(pairs)[1]) { 
  temp_meth <- t(methylation2[pairs$probe[i],])
  temp_exp <- t(expression2[pairs$gene_symbol[i],])
  
  correlation <- cor.test(temp_meth[,1], temp_exp[,1], method = c("spearman"),adjust.method="fdr",exact=FALSE)
  
  overall_results$p.val[i] <- correlation$p.value
  overall_results$rho[i] <- correlation$estimate
  overall_results$statistic[i] <- correlation$statistic
  
}
```

Keep only significant results (nominal pval < 0.05)
```{r}
eqtms <-overall_results[which(overall_results$p.val < 0.05),]
```

Save significant eQTMs 
```{r}
write.csv(eqtms, "maryland_AFR-sig_eQTMs_pval.csv", row.names = F)
```