---
title: "Run eQTM analysis on expression and methylation residuals"
author: "Kyriaki Founta"
output: html_document
date: "2025-01-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("readr")
library("stats")
```

# Load data
Probe-gene pairs, along with genomic coordinates
```{r}
probe_gene_pairs <- read_csv("probe-gene-pairs.csv")
View(probe_gene_pairs)
```
Gene expression residuals
```{r}
expression <- read_csv("residuals_expression_admix_proportions.csv")
head(expression)
```

Beta value residuals
```{r}
methylation <- read_csv("residuals_methylation_admix_proportions.csv")
head(methylation)
```
Keep same patients between methylation and expression matrix
```{r}
colnames(methylation)[1] <- "probe"
methylation <- methylation[,c("probe",intersect(colnames(methylation), colnames(expression)))]
```

Create appropriate format for beta matrix
```{r}
methylation <- as.data.frame(methylation)
row.names(methylation) <- methylation[,1]
methylation <- methylation[,-1]
```
Preprocess and format expression data appropriately
```{r}
expression <- as.data.frame(expression)

row.names(expression) <- expression[,1]
expression <- expression[,-1]
```
From file with candidate eQTMs, keep only probe-gene pairs
```{r}
pairs <- probe_gene_pairs[,c(6,12)]
colnames(pairs) <- c("probe", "gene_symbol")
```

## Prepare to run the eQTM analysis
Create dataframe to store the results
```{r}
overall_results <- data.frame(probe = pairs$probe, gene=pairs$gene_symbol)
```

Write loop
```{r}
for (i in 1:dim(pairs)[1]) { 
  temp_meth <- t(methylation[pairs$probe[i],])
  temp_exp <- t(expression[pairs$gene_symbol[i],])
  
  correlation <- cor.test(temp_meth[,1], temp_exp[,1], method = c("spearman"),adjust.method="fdr",exact=FALSE)
  
  overall_results$p.val[i] <- correlation$p.value
  overall_results$rho[i] <- correlation$estimate
  overall_results$statistic[i] <- correlation$statistic
  
}
```

Correct for multiple testing
```{r}
overall_results$p.adjusted <- p.adjust(overall_results$p.val, method = "BH")
```

Keep only results with adjusted pval < 0.05
```{r}
eqtms <-overall_results[which(overall_results$p.adjusted < 0.05),]
dim(eqtms)
```

Save significant eQTMs 
```{r}
write.csv(eqtms, "p.adjust_sig_eQTMs_residuals_admix_proportions.csv", row.names = F)
```
