---
title: "Mediation Analysis for HLA community meQTL/eQTMs"
author: "Kyriaki Founta"
date: "2026-04-17"
output: html_document
---

```{r}
library(readr)
library(mediation)  
library(tidyverse)
library(dplyr)
```

## Load HLA eQTM community
Load a list with all probes and genes in the HLA eQTM community
```{r}
comm <- read_csv("HLA_eQTM_community_networkx.txt",
                 col_names = FALSE)
```

## Load data
Load all significant eQTMs
```{r}
eqtms <- read_csv("p.adjust_sig_eQTMs_residuals_admix_proportions.csv")
head(eqtms)
```
Load gene expression residuals
```{r}
expression <- read_csv("residuals_expression_admix_proportions.csv")
head(expression)
```

Load methylation residuals
```{r}
methylation <- read_csv("residuals_methylation_admix_proportions.csv")
head(methylation)
```
Keep same patients between methylation and expression matrix
```{r}
colnames(methylation)[1] <- "probe"
methylation <- methylation[,c("probe",intersect(colnames(methylation), colnames(expression)))]
```

Create appropriate format for beta value matrix
```{r}
methylation <- as.data.frame(methylation)
row.names(methylation) <- methylation[,1]
methylation <- methylation[,-1]
head(methylation)
```

Preprocess expression data
```{r}
expression <- as.data.frame(expression)

row.names(expression) <- expression[,1]
expression <- expression[,-1]

head(expression)
```

From file with candidate eQTMs, keep only probe-gene pairs
```{r}
pairs <- eqtms[,c(1,2)]
colnames(pairs) <- c("probe", "gene")
head(pairs)
```

Now only select probes and genes in HLA community
```{r}
pairs2 <- pairs[which(pairs$probe %in% comm[,1]),]
pairs3 <- pairs2[which(pairs2$probe %in% comm[,1]),]
dim(pairs3)
```

## meSNP Genotypes
Load genotypes for meSNPs obtained from the TCGA BRCA .vcf files (first column: variant id (meSNP), second column: probe (meQTL targeted probe), the rest of the columns: genotypes per variant per sample)
```{r}
gts <- read_csv("GTs_hla_eqtms.csv")

colnames(gts)[2] <- "probe"

head(gts)
```
Process GT df
```{r}
gts2 <- as.data.frame(unique(gts[,-2]))

row.names(gts2) <- gts2[,1]
gts2 <- gts2[,-1]

head(gts2)
```
Keep same samples across datasets
```{r}
meth <- methylation[,intersect(colnames(methylation), colnames(gts2))]
gts3 <- gts2[,colnames(meth)]
expr <- expression[,colnames(meth)]

# Sanity check No 1
all.equal(colnames(meth), colnames(gts3))
```

```{r}
# Sanity check No 2
all.equal(colnames(meth), colnames(expr))
```

## Create dataframe to run mediation analysis on 
Add GTs per variant in variant-probe pairs df
```{r}
nets <- merge(pairs3, gts[,c(1:2)], by = "probe")
head(nets)
```

## Run mediation analysis with boot = FALSE
Set up null dataframe
```{r}
library(mediation)

overall_results <- data.frame(  
  probe = nets$probe,  
  gene = nets$gene,  
  variant_id = nets$variant_id,  
  ACME_estimate = NA,  
  ACME_ci_lower = NA,  
  ACME_ci_upper = NA,  
  ACME_pval = NA,  
  ADE_estimate = NA,  
  ADE_ci_lower = NA,  
  ADE_ci_upper = NA,  
  ADE_pval = NA,  
  total_effect_estimate = NA,  
  total_effect_pval = NA,  
  prop_mediated_estimate = NA,  
  prop_mediated_pval = NA  
)  
```

Run mediation analysis in a loop
```{r}
n_sims <- 5000  

for (i in 1:nrow(nets)) {  
    
  # Extract data for this triplet  
  snp_name   <- nets$variant_id[i]  
  probe_name <- nets$probe[i]  
  gene_name  <- nets$gene[i]  
    
  temp_geno <- as.numeric(gts3[snp_name, ])  
  temp_meth <- as.numeric(meth[probe_name, ])  
  temp_expr <- as.numeric(expr[gene_name, ])  
    
  # Build temporary data frame  
  temp_df <- data.frame(  
    genotype = temp_geno,  
    mediator = temp_meth,  
    outcome  = temp_expr  
  )  
    
  # Remove rows with missing values  
  temp_df <- na.omit(temp_df)  
    
  tryCatch({  
    # Mediator model: Methylation ~ SNP  
    model_m <- lm(mediator ~ genotype, data = temp_df)  
      
    # Outcome model: Expression ~ SNP + Methylation  
    model_y <- lm(outcome ~ genotype + mediator, data = temp_df)  
      
    # Mediation analysis  
    med_result <- mediate(model_m, model_y,  
                          treat    = "genotype",  
                          mediator = "mediator",  
                          boot     = FALSE,  
                          sims     = n_sims)  
      
    med_summary <- summary(med_result)  
      
    # ACME = Average Causal Mediation Effect (indirect: SNP → Meth → Expr)  
    overall_results$ACME_estimate[i]  <- med_summary$d0  
    overall_results$ACME_ci_lower[i]  <- med_summary$d0.ci[1]  
    overall_results$ACME_ci_upper[i]  <- med_summary$d0.ci[2]  
    overall_results$ACME_pval[i]      <- med_summary$d0.p  
      
    # ADE = Average Direct Effect (direct: SNP → Expr, bypassing Meth)  
    overall_results$ADE_estimate[i]   <- med_summary$z0  
    overall_results$ADE_ci_lower[i]   <- med_summary$z0.ci[1]  
    overall_results$ADE_ci_upper[i]   <- med_summary$z0.ci[2]  
    overall_results$ADE_pval[i]       <- med_summary$z0.p  
      
    # Total effect  
    overall_results$total_effect_estimate[i] <- med_summary$tau.coef  
    overall_results$total_effect_pval[i]     <- med_summary$tau.p  
      
    # Proportion mediated  
    overall_results$prop_mediated_estimate[i] <- med_summary$n0  
    overall_results$prop_mediated_pval[i]     <- med_summary$n0.p  
      
  }, error = function(e) {  
    message(paste("Error at triplet", i,   
                  "(", snp_name, probe_name, gene_name, "):", e$message))  
  })  
    
  # Optional: print progress  
  if (i %% 100 == 0) message(paste("Completed", i, "of", nrow(nets)))  
}  
```

Correct for multiple testing
```{r}
overall_results$ACME_p_adjusted <- p.adjust(overall_results$ACME_pval, method = "BH")  
overall_results$ADE_p_adjusted  <- p.adjust(overall_results$ADE_pval, method = "BH")  
overall_results$total_effect_p_adjusted  <- p.adjust(overall_results$total_effect_pval, method = "BH")  
overall_results$prop_mediated_p_adjusted <- p.adjust(overall_results$prop_mediated_pval, method = "BH")  
```

Examine significant results
```{r}
# Triplets where methylation significantly mediates the SNP→Expression effect  
significant_mediations <- overall_results[which(overall_results$ACME_p_adjusted < 0.05), ]  
message(paste(nrow(significant_mediations), "significant mediations out of", nrow(overall_results)))

head(significant_mediations[order(significant_mediations$ACME_p_adjusted), ])  
```
Save significant mediation results
```{r}
write.csv(significant_mediations[order(significant_mediations$ACME_p_adjusted), ], "sig_mediation_5000_sims.csv",
          row.names = F)
```
