---
title: "Differential Methylation Analysis in the NCI-Maryland Cohort"
author: "Kyriaki Founta"
date: "2026-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data
Libraries
```{r}
library(readr)
library(limma)
library(dplyr)
library(stringr)
library(readxl)
```
Load methylation metadata downloaded from GEO
```{r}
meth_metadata <- read_csv("meth_metadata.csv")
```

Subset columns of interest and clean-up necessary information
```{r}
sel <- meth_metadata[,c("title","geo_accession","race:ch1","molecular subtype:ch1", "age_at_surgery:ch1","Sex:ch1","ndi continuous:ch1")]

sel$accession <- str_sub(sel$title,32,36)
sel$tissue_type <- word(sel$title, -1) 

colnames(sel)[c(4,5,6,7)] <- c("subtype", "age", "sex", "ndi")

head(sel)
```

Load genetic ancestry admixture proportions
```{r}
ancestry <- read.csv("final_ancestry_maryland.csv")
```
Load estimated tumor purity
```{r}
purity <- read.csv("normal-mast_purity_InfPurify.csv")
```

```{r}
colnames(purity) <- c("accession", "purity")
```

Combine all data
```{r}
merged <- merge(sel[which(sel$tissue_type == "tumor"),c("accession","subtype","age","sex")], ancestry, by ="accession")
```
Add purity
```{r}
final <- unique(merge(merged, purity, by ="accession"))
```

Create categorical ancestry anntoation based on 80% admixture threshold
```{r}
final <- final %>%  
  mutate(ancestry = case_when(  
    AFR > 0.8 ~ "AFR",  
    EUR > 0.8 ~ "EUR",  
    AMR > 0.8 ~ "AMR",  
    EAS > 0.8 ~ "EAS",  
    SAS > 0.8 ~ "SAS",  
    TRUE ~ "Admixed"  
  ))

table(final$ancestry)
```
Remove male individuals
```{r}
final2 <- final[which(final$sex != "M"),]
```

```{r}
row.names(final2) <- final2$accession
final2 <- final2[,-1]
```

Remove samples with missing demographic values
```{r}
final4 <- na.omit(final2) 
```
Load tumor beta values
```{r}
betas <- read_csv("tumors_enmix_betas.csv")
head(betas)
```
Process tumor methylation
```{r}
colnames(betas) <- betas[1,]
betas <- betas[-1,]
```

```{r}
betas <- as.data.frame(betas)

row.names(betas) <- betas[,1]
betas <- betas[,-1]
```

Keep only samples with metadata available
```{r}
mat <- as.matrix(betas[,row.names(final4)])
all.equal(colnames(mat), row.names(final4)) # sanity check for sample order
```

Trasnform beta values to matrix
```{r}
mval <- as.matrix(mat)
```

Make subtype column compatible with limma syntax
```{r}
final4[which(final4$subtype == "HR+"),"subtype"] <- "HRpos"
final4[which(final4$subtype == "HER+"),"subtype"] <- "HERpos"
```

## Filter out EPIC-450K non-shared probes
Load and process EPIC metadata file
```{r}
epic <- read_csv("infinium-methylationepic-v-1-0-b5-manifest-file.csv", skip = 7)

epic2 <- as.data.frame(epic)

row.names(epic2) <- epic2[,1]
epic2 <- epic2[,-c(1:2)]

epic2$Methyl450_Loci[which(is.na(epic2$Methyl450_Loci))] <- "FALSE"
```
Subset epic data to 450k array shared probes
```{r}
shared <- epic2[which(epic2$Methyl450_Loci == "TRUE"),]
```
Keep only EPIC array probes shared with 450K array (to enable comparison with the TCGA BRCA results)
```{r}
mval2 <- mval[which(row.names(mval) %in% row.names(shared)),]
```

## Run differential methylation analysis
First, design the model
```{r}
design_full <- model.matrix(~0 + AFR + purity + age + subtype, data=final4) 
head(design_full)
```
Fit the model
```{r}
fit <- lmFit(mval2, design_full)
```

Fit contrasts and recalibrate model
```{r}
fit2 <- eBayes(fit)
```

Final run the differential methylation function
```{r}
DMPs_subtype <- topTable(fit2, num=Inf, coef="AFR",  adjust.method = "BH") 
head(DMPs_subtype)
```
Get significant results based on q-value
```{r}
sig.probes <- DMPs_subtype[DMPs_subtype$adj.P.Val < 0.05,]
dim(sig.probes)
```

Further filter out results with low effect sizes
```{r}
effective <- sig.probes[abs(sig.probes$logFC) >= 0.1,]
dim(effective)
```
Save results
```{r}
write.csv(effective,"AFR-significant_probes_revised-meta_450k-only.csv", row.names = T)
```

